---
title: "Lines - tracking trends"
subtitle: "Session 5 practical exercises"
editor_options: 
  chunk_output_type: inline
---

```{r}
#| include: false
library(webexercises)
library(pacman)
p_load(rio, here, tidyverse)
Sys.setlocale("LC_TIME", "English")

imd <- import(here("data", "clean", "IMD_Sample_Clean.rds"))
```

Barplots are useful for categorical counts and combinations of those. When we account for four years - that is: four data points, the resulting graph can be adequate. Increase that number of data points, and it becomes increasingly difficult to read as well, so let's start moving towards finer time scales and learn about useful geoms to assist us in our tasks.

Kassandra's next plot request is about case trends overall and by sex. She is interested in learning more about disease seasonality and asks us to deliver a monthly trend line with the three categories.

Continue in your Session 5 script, where your data and libraries are already loaded.

## Part 8 · Lines and Trends

### Too many data points, the wrong way

**Action** — Count cases by `key_date` with `count()`, then plot `x = key_date`, `y = n` with `geom_line()`.

Technically nothing is wrong here — you counted first, exactly like you are supposed to. And yet the result is a saw-toothed mess, close to unreadable. The problem is not that you forgot to aggregate, it is that you aggregated at the **wrong resolution**. Most single days have zero, one or two cases, and that day-to-day noise buries any real pattern underneath it.

::: callout-note
## Choosing a time unit is not optional

A line chart needs one value per position on `x`, but *which* position is a choice you make, not something the data hands you. Four years of daily counts is too fine to read; four years of yearly counts (four points total) would be too coarse to say anything about seasonality. Monthly sits in between, and it is the resolution Kassandra actually needs to talk about seasonality.
:::

### Floor the date to build a monthly series

You already filtered out unknown `sex` values in earlier sessions when a table needed complete groups. We will do the same here, and be explicit about it: **cases with an unrecorded sex are excluded from both the total and the by-sex lines**, so that the two add up to exactly the same numbers when you compare them. Again, it's an important decision to make at the institutional level when working on reports.

We can have daily counts thanks to `key_date`. We need `lubridate` to use that date to incorporate `month` variable in the data to count. As usual, many function could assist us, but this is the one we will be using today: `floor_date()`. It has two main arguments:

1.  A `<date>` vector as first argument. Within a tidy pipe, a date variable

2.  The `unit = " "` where you write the desired grouping, for example "week", "month", or "year". If you also had time (h:m:s) in the date, you could use "hour", or "minute"

The function takes the date and **rounds it down** to its nearest "floor" date. If we want months, that means setting all cases happening at any day of January 1999 to "1999-01-01", the "floor date" of that month. If the unit is weeks, it does the same with every week's Monday or Sunday (EU vs USA).

Since the output is still a real date, you get naturally ordered categories for free. More convenient, impossible.

**Action** — Create `imd_monthly_sex`: filter out missing `sex`, create a `month` column with `floor_date(key_date, unit = "month")`, then `count(month, sex)`.

**Action** — Now create `imd_monthly_total` the same way, but without `sex` in `count()` — just `count(month)`. Remember to filter out missing `sex` here too, so both tables describe exactly the same set of cases.

### A trick: turn a summary into a third category

You now have two separate tables that both describe cases over time — one split by sex, one not. To plot them as three lines on the same chart with one shared legend, the cleanest way is to make "Total" look like a third value of `sex`, and stack both tables into one using `bind_rows().`

`bind_rows()` is a Tidy function that takes dataframes as arguments, and paste their rows together. There is one condition, though: they must have exactly the **same variable in number, name and class**. <!--# not really true, i.e. if we bind them, sex for total will just have NA, with no warning or error --> And we are missing one variable in the total count:

**Action** — Add a new column to `imd_monthly_total`: `sex = "Total"` — a fixed value, the same for every row, not derived from anything. Identical name and class.

**Action** — Combine both tables with `bind_rows()` into `imd_monthly_sex_total`.

Check `imd_monthly_sex_total` in your viewer. `sex` now has three values instead of two — `"Male"`, `"Female"`, and your fabricated `"Total"` — sitting in a single tidy column, ready for a single `aes(color`

`= sex)`.

**Action** — Plot your case trend with `month` on the `x` axis and number of cases in the `y`. Map the color to `sex` to have separate colors and the legend

### Manual palette: choosing on your own

If we want custom colors, `scale_color_brewer()` would work here, but Brewer palettes assign colors automatically in a fixed order — you cannot tell it "Total must be black". When a specific category needs a specific, deliberate color, you write the mapping yourself with `scale_color_manual()`.

It takes more time and effort compared with predefined palettes, but it's worth taking a moment to learn this important action. The function will use a single argument today: `values`. Inside, we will open a vector `c()` and include pairs of `level = color` arguments separated by commas, making sure all our categories are included.

This time, the code is on the house

**Action** — Add `scale_color_manual()`, with `values` set to a named vector including our three categories of `sex`. You can choose your colors, of course.

::: {.callout-tip collapse="true" appearance="simple" icon="false"}
## Manual scale

``` r
scale_color_manual(
  values = c(
    "Total"  = "black",
    "Male"   = "steelblue",
    "Female" = "firebrick"
  )
)
```
:::

Notice, that when we were colouring shapes (bars or columns), we used scale fill function. Now that we're colouring lines, we need to use scale color function.

## Final plot - Trend by sex

Finnish the plot:

**Action** — Add `scale_x_date()` with 4-months `date_breaks` and `date_labels` in month-year with `"%b %Y"`.

**Action** — Add a theme of your choice, and `labs()` for the axes, legend title and a report-ready title.

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 8.5
#| out-width: 80%
# Monthly series by sex (unknown sex excluded)
imd_monthly_sex <- imd %>%
  filter(!is.na(sex)) %>%
  mutate(
    month = floor_date(key_date, unit = "month")
  ) %>%
  count(month, sex)

# Monthly total (same exclusion, so totals match Male + Female)
imd_monthly_total <- imd %>%
  filter(!is.na(sex)) %>%
  mutate(
    month = floor_date(key_date, unit = "month")
  ) %>%
  count(month) %>%
  mutate(
    sex = "Total"
  )

# The trick: stack both tables into one
imd_monthly_sex_total <- bind_rows(
  imd_monthly_sex,
  imd_monthly_total
)

imd_monthly_sex_total %>%
  ggplot(aes(
    x = month,
    y = n,
    color = sex
  )) +
  geom_line(
    linewidth = 0.8
  ) +
  scale_color_manual(
    values = c(
      "Total"  = "grey20",
      "Male"   = "steelblue",
      "Female" = "darkorange"
    )
  ) +
  scale_x_date(
    date_breaks = "4 months",
    date_labels = "%b %Y"
  ) +
  labs(
    x = "Month",
    y = "Number of cases",
    color = "Sex",
    title = "Monthly IMD cases, overall and by sex"
  ) +
  theme_classic() 
```

## Part 9 · From line to area

### A line is a bar in disguise

Think back to `geom_bar()`/`geom_col()` versus `geom_line()`. Both take the same kind of input — one value per position — and simply draw it differently: as a bar, or as a connected point. `geom_area()` is a third way of drawing that same idea: a line, with the space underneath it filled in.

Recover the code of the final plot of serogroup cases by year. We will turn it into an area plot of monthly cases, using `floor_date()` again. Copy this code to see for yourself, but don't spend any more time on this - you still have exercises to do!

```{r}
#| fig-align: "center"
#| out-width: "80%"
# Monthly series by serogroup
imd_monthly_serogroup <- imd %>%
  mutate(
    month = floor_date(key_date, unit = "month")
  ) %>%
  count(month, serogroup)

# Final plot (2): stacked area by serogroup
serogroup_area <- imd_monthly_serogroup %>%
  ggplot(aes(
    x = month,
    y = n,
    fill = serogroup
  )) +
  geom_area() +
  scale_fill_brewer(
    palette = "Set2"
  ) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  labs(
    x = "Month",
    y = "Number of cases",
    fill = "Serogroup",
    title = "Monthly IMD cases by serogroup"
  ) +
  theme_classic()

serogroup_area
```

By default, `geom_area()` stacks — each serogroup's monthly count sits on top of the previous one, so the height of the whole stack at any given month is the national total for that month. Instead of seven crossing lines, you get one continuous silhouette, split into colored bands. The best of the bar and the best of the line.

------------------------------------------------------------------------

## Exercise Summary

Two graphs, two different problems solved. The first taught you a genuinely reusable trick: turning a separate summary into a fabricated category of the same variable, so `bind_rows()` and a single `aes()` can put it on equal footing with the real categories and, from there, `scale_color_manual()` to give a meaningful category its own deliberate color.

The second showed that bar, line and area are three faces of the same idea, and that switching from line to area is often the fix when too many crossing lines stop being readable. In the next excercise, the calendar itself becomes the subject of the plot, one day at a time.

| Function | Package | What it does |
|------------------------|------------------------|------------------------|
| `floor_date()` | lubridate | Rounds a date down to the start of a given unit (day, month, year...) |
| `bind_rows()` | dplyr | Stacks two tables with matching columns into one |
| `scale_color_manual()` | ggplot2 | Assigns specific colors to specific category values, written by hand |
| `geom_area()` | ggplot2 | Like `geom_line()`, with the area underneath filled; stacks by default when `fill` is mapped |

::: {.callout-tip collapse="true"}
## 💡 Show solution — only after trying yourself!

``` r
# Load libraries
library(pacman)
p_load(rio, here, tidyverse)

# Import data
imd <- import(here("data", "clean", "IMD_Sample_Clean.rds"))

# The trap: daily resolution is too fine
imd %>%
  count(key_date) %>%
  ggplot(aes(
    x = key_date,
    y = n
  )) +
  geom_line()

# Monthly series by sex (unknown sex excluded)
imd_monthly_sex <- imd %>%
  filter(!is.na(sex)) %>%
  mutate(
    month = floor_date(key_date, unit = "month")
  ) %>%
  count(month, sex)

# Monthly total (same exclusion, so totals match Male + Female)
imd_monthly_total <- imd %>%
  filter(!is.na(sex)) %>%
  mutate(
    month = floor_date(key_date, unit = "month")
  ) %>%
  count(month) %>%
  mutate(
    sex = "Total"
  )

# The trick: stack both tables into one
imd_monthly_sex_total <- bind_rows(
  imd_monthly_sex,
  imd_monthly_total
)

# Final plot (1): trend by sex, with a manual color scale
trend_by_sex <- imd_monthly_sex_total %>%
  ggplot(aes(
    x = month,
    y = n,
    color = sex
  )) +
  geom_line(
    linewidth = 0.8
  ) +
  scale_color_manual(
    values = c(
      "Total"  = "grey20",
      "Male"   = "steelblue",
      "Female" = "darkorange"
    )
  ) +
  scale_x_date(
    date_breaks = "4 months",
    date_labels = "%b %Y"
  ) +
  labs(
    x = "Month",
    y = "Number of cases",
    color = "Sex",
    title = "Monthly IMD cases, overall and by sex"
  ) +
  theme_classic()

# Save the plot
ggsave(
  plot     = trend_by_sex,
  filename = "monthly_trend_sex.png",
  path     = here("output"),
  units    = "in",
  width    = 7,
  height   = 5,
  dpi      = 300
)

# Monthly series by serogroup
imd_monthly_serogroup <- imd %>%
  mutate(
    month = floor_date(key_date, unit = "month")
  ) %>%
  count(month, serogroup)

# Final plot (2): stacked area by serogroup
serogroup_area <- imd_monthly_serogroup %>%
  ggplot(aes(
    x = month,
    y = n,
    fill = serogroup
  )) +
  geom_area() +
  scale_fill_brewer(
    palette = "Set2"
  ) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  labs(
    x = "Month",
    y = "Number of cases",
    fill = "Serogroup",
    title = "Monthly IMD cases by serogroup"
  ) +
  theme_classic()
```
:::

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function() {
  var radiogroups = document.getElementsByClassName("webex-radiogroup");
  for (var i = 0; i < radiogroups.length; i++) {
    radiogroups[i].onchange = radiogroups_func;
  }
});
</script>
```
