---
title: "Histograms for Epicurves"
subtitle: "Session 5 practical exercises"
---

```{r}
#| include: false
library(webexercises)
library(pacman)
p_load(rio, here, tidyverse, aweek)
Sys.setlocale("LC_TIME", "English")

imd <- import(here("data", "clean", "IMD_Sample_Clean.rds"))
imd_2000 <- imd %>%
  filter(year == 2000) %>% 
  mutate(
    week_date = floor_date(key_date, unit = "week", week_start = 1)
  )
```

Kassandra's report has one section left. She wants to zoom-in on the peak year of cases, `2000`, to understand its trend and look for relevant changes there. We saw that in the tables produced in Session 4. She needs an epicurve of weekly cases for the year 2000.

Continue in your Session 5 script, where your data and libraries are already loaded. We are adding one new package today, `aweek`, just for a small comparison later on.

## Part 10 · Epicurves: where time is the protagonist

Every plot so far has counted something categorical — a serogroup, a sex, a year or month. Today the categories do not exist yet. You have one raw date per case, and from there we need to build our own bins. In the previous exercise, we used `floor_date()` to get months categories with dates for the line plots.

An epicurve is tipically focused around a short period of time: days or weeks (outbreaks), weeks or months (long outbreaks, epi seasons) or a year (surveillance). And usually, even lengthy series with months or years of data get grouped around epidemiological weeks. This level of granularity implies other aspects in terms of grouping and manipulating dates.

Let's start by exploring the different aspects of weeks.

### From weeks to epiweeks

**Action** — Create a new dataset with only cases from the year 2000, name it `imd_2000`.

**Action** — Create a histogram of cases using `key_date` as `x` variable, without any further argument

It works, no error this time — but why does a date happily sit on a continuous axis and get chopped into bins at all? Because underneath, a date **is** a number: the count of days since 1970-01-01. So whatever `binwidth` you set with a date variable, you are setting it in **days**.

You will also notice a console message, something like `stat_bin() using bins = 30. Pick better value with binwidth`. R is telling you that the width of the bin is your decision, not its.

::: callout-note
## R being honest with you (again)

`geom_histogram()` has to draw *something* even if you never tell it how wide a bin should be, so it silently picks 30 equal-width bins across your data and warns you about it. Thirty is not a magic number that fits every dataset — it is a placeholder, and the message is R's way of refusing to let that placeholder pass unnoticed.
:::

**Action** — Try `binwidth = 1`, then `7`, then `30`. Watch how the same cases tell a completely different story depending on the width you choose.

### Which week is it, exactly?

`binwidth = 7` finally starts looking like a real epidemic curve — one bar per week. But which day did R choose to start counting from? In `binwidth = 7` what happens with 31-days months, or February? R picks the origin of the first bin on its own, silently, without any actual calendar or epidemiological criteria.

`lubridate` is our date package. Period. But within it, many options are available. And we have several candidate functions for getting the job done: `year()` and `week()` together with `isoyear()` and `isoweek()`. So what is the difference?

When working in epiweeks, the usual label we see is "2000-W01" meaning first epiweek of the year 2000. Should be easy, right?

**Action** — In your `imd_2000` data, add a mutate to create the `year` `isoyear`, `week` and `isoweek` variables using the lubridate functions.

**Action** — Open the viewer, arrange the data by `key_date` and check the newly created columns.

**Action** — Run this small comparison and look closely at the columns:

```{r}
#| eval: false
tibble(
  key_date = as.Date(c("1999-12-31", "2000-01-01", "2000-01-02")),
  year     = year(key_date),
  isoyear  = isoyear(key_date),
  isoweek  = isoweek(key_date),
  week     = week(key_date)
)
```

All three dates share the exact same `isoweek`, 52 — they are the same ISO week, spanning the turn of the year. But `year()` reads the calendar year printed on the date, so two of these rows say "2000" while `isoyear()` correctly keeps all three in 1999, the year that ISO week 52 actually belongs to. The same happens with `isoweek()` and `week()`.

::: callout-note
## R being honest with you, calendar edition

Pair `isoweek()` with the wrong year function and you invent a "2000-W52" that never happened, while the real ISO week 52 gets torn into two pieces — a few cases wrongly reported under next year's week 1. This is not a hypothetical: it is one of the most common labelling errors in published outbreak reports.
:::

::: callout-caution
## If the function `epiweek()` exist, why not use it?

`lubridate::epiweek()` looks like the obvious shortcut, but it follows the **MMWR/CDC** definition (weeks start on Sunday), not the **ISO** definition (weeks start on Monday) that EPIET and ECDC use. Same dates, different week numbers. Always `isoweek()` here.
:::

### The right epiweek label

Now that you know *why* the year matters, let's fix the x axis itself. `isoweek()` produced only an ordered sequence of numbers that we could use as a continuous scale, right? But `dates` enable many more customizing options, so how do we keep the best of both?

`floor_date()` — already familiar from the previous exercise — takes a `unit = "week"` argument, and with `week_start = 1` it rounds every date down to the **Monday** that starts its week. A real date, ready to sit on a continuous axis, gaps and all. It's the same logic of how isoweeks are defined, to begin with.

**Action** — Create a `week_date` variable in `imd_2000` using `floor_date()` with the appropriate arguments. Does `week_date` and `isoweek` match 100%? how would you check that?

::: callout-caution
## A small edge effect

Because `imd_2000` was already filtered to `year == 2000`, the very first and last weeks of your table may look slightly short — a case on 1 January could belong to a week that started in late December 1999, and it is missing from this table. For this exercise we accept that small clipping at the edges; a production report would filter on the ISO week itself, not the calendar year.
:::

**Action** — Build two histograms, one using `week_date` and other with `isoweek`

-   Which bindwith should you choose for each one?

-   Do you spot any difference?

::: {.callout-tip collapse="true" appearance="simple" icon="false"}
## Hint

For `week_date`, it should be 7 because it's a date variable, and dates come in days by default - even if it is only a grouping, non-contiguous date.

For `isoweek`, it should be 1 because it is read as a simple numeric variable, and we want every number represented in the plot

The difference lies in the first and last bars: with `week_date` the first bar is the 6 cases happening the last isoweek 1999, and is ordered. With `isoweek` those were merged with the isoweek 52 of 2000, creating a larger final column
:::

::: callout-warning
## The devil is in the details

Why bother so much with isoweeks and correct counts? The simple histogram from the very beginning would have sufficed, nobody would have noticed. When you get decent results at first try, it is tempting to take them for granted. Truth is, the first counts were not correct. We need to create good graphs - and also reliable ones.
:::

### Aweek for direct label creation

Before moving on, a quick look at what `aweek` library with the function`date2week()` that accounts for all we mentioned, and produces directly nice labels ready to plot - even if we need this wide set of arguments for it:

```{r}
#| eval: false
date2week(
  as.Date(c("1999-12-31", "2000-01-01", "2000-01-02")),
  week_start = "Monday",   # ISO/ECDC standard
  floor_day = T,           # Don't include day of the week in the label
  factor = T               # For a ready-to-use, ordered factor
)
```

This returns the label directly, `"1999-W52"` for all three dates — the same answer as your `isoyear`/`isoweek` columns above, just already glued together. Then, you could just use `geom_col()` to create the epicurve, but loosing in the way all the date customization available. You will see why:

## Part 11 · Labelling the calendar

Your weekly histogram with `week_date()` has correct spacing and correct grouping — the hard part is done. What is left is dressing the axis so it reads like a report, not like a debugging session.

**Action** — Add `scale_x_date()` with `date_breaks = "1 month"` and `date_labels = "%b"`, the `strftime` codes you already used earlier in the session.

**Action** — Add `theme_minimal()`

That looks clean at monthly resolution. But we did not build a whole session on epiweeks for nothing: grouping aside (which is super important) we also want nice and publication-ready labels that reflect the weeks nicely. `strftime` codes got your back:

**Action** — Change `date_breaks` to `"2 weeks"`, and change `date_labels = "%g-W%V"`. Run it.

The labels start overlapping into an unreadable smudge, we can't even tell how the labels look like! Twice as many ticks, same amount of horizontal space. This is the first time you need `theme()` to do more than pick a predefined style.

**Action** — Just copy the code and add it:

::: {.callout-tip collapse="true" appearance="simple" icon="false"}
## Angled x axis labels

``` r
theme_minimal() +
theme(
  axis.text.x = element_text(
    angle = 45,
    hjust = 1
  )
)
```
:::

Legible again — but look at what the ticks actually say: `"1999-W51"`, `"2000-W01"`, isoyear and week labels. Kassandra's report does not talk in calendar dates, it talks in ISO weeks, exactly the numbers you built by hand in Part 10.

::: callout-tip
## Save the code for later

At the end of every ggplot exercise we are giving you pieces of "extra" code: more advanced tricks outside the scope of this course. Why bother in doing so? So you can have it and use it later. Even if you don't understand how `strftime` or `theme()` works, you have two incredibly powerful lines of code that you will recycle in the future.
:::

## Final plot

**Action** — Finish the plot: add some color, or `aes()` if you feel like it and have time. Otherwise, here is a nice final version of it:

```{r}
#| fig-align: "center"
#| fig-width: 8
#| out-width: "80%"
# Final plot
imd_2000 %>% 
  ggplot(aes(x = week_date)) +
  geom_histogram(binwidth = 7, 
                 color = "navy",
                 fill = "steelblue") +
  scale_x_date(
    date_breaks = "2 weeks",
    date_labels = "%G-W%V"
  ) +
  labs(
    x = "Epidemiological week",
    y = "Number of cases"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

------------------------------------------------------------------------

## Exercise Summary

You built the same weekly epicurve several times over, and each version quietly fixed something the previous one got wrong. A bare `geom_histogram()` on `key_date` showed you that a date is just a number underneath, and `binwidth` gave you control over the story those numbers tell.

Then came the real trap: natural weeks on their own are not the same than epidemiological ISO weeks, the kind of error that slips into a published surveillance report unnoticed. `isoweek()` and `floor_date()` with `week_start = 1` gave you a real start of the week to anchor the x axis on, and `%G`/`%V` finally let the labels speak in ISO weeks without giving up the correctly-spaced calendar underneath.

Finally, a new trick for the `theme()` function that will come in handy many times when dealing with either very long or numerous x labels that overlap, downgrading the appearance of your plot.

| Function | Package | What it does |
|------------------------|------------------------|------------------------|
| `geom_histogram()` | ggplot2 | Counts rows into bins along a continuous axis; needs an explicit `binwidth` |
| `binwidth` | ggplot2 | Width of each bin, in the units of the mapped variable — days, for a `Date` |
| `year()`, `week()` | lubridate | Calendar year and lubridate's own week-of-year count — not ISO |
| `isoyear()`, `isoweek()` | lubridate | ISO 8601 year and week number; always used as a pair |
| `epiweek()` | lubridate | MMWR/CDC week definition (Sunday-start) — not ISO, not used here |
| `floor_date(unit = "week", week_start = 1)` | lubridate | Rounds a date down to the Monday that starts its ISO week |
| `date2week()` | aweek | Converts dates directly into ready-made ISO week labels |
| `%V`, `%G` | base R (strftime) | ISO week number and its corresponding ISO year, usable inside `date_labels` |
| `theme(axis.text.x = element_text(angle, hjust))` | ggplot2 | Rotates axis tick labels so they stop overlapping |

::: {.callout-tip collapse="true"}
## 💡 Show solution — only after trying yourself!

``` r
# Load libraries
library(pacman)
p_load(rio, here, tidyverse, aweek)

# Import data
imd <- import(here("data", "clean", "IMD_Sample_Clean.rds"))

# Filter to the peak year
imd_2000 <- imd %>%
  filter(year == 2000)

# Crude histogram, no arguments
imd_2000 %>%
  ggplot(aes(
    x = key_date
  )) +
  geom_histogram()

# Same plot at three bin widths
imd_2000 %>%
  ggplot(aes(
    x = key_date
  )) +
  geom_histogram(binwidth = 1)

imd_2000 %>%
  ggplot(aes(
    x = key_date
  )) +
  geom_histogram(binwidth = 7)

imd_2000 %>%
  ggplot(aes(
    x = key_date
  )) +
  geom_histogram(binwidth = 30)

# Compare every year/week candidate side by side
imd_2000 <- imd_2000 %>%
  mutate(
    year      = year(key_date),
    isoyear   = isoyear(key_date),
    week      = week(key_date),
    isoweek   = isoweek(key_date),
    week_date = floor_date(key_date, unit = "week", week_start = 1)
  )

imd_2000 %>%
  arrange(key_date) %>%
  select(key_date, year, isoyear, week, isoweek, week_date)

# The trap, made explicit
tibble(
  key_date = as.Date(c("1999-12-31", "2000-01-01", "2000-01-02")),
  year     = year(key_date),
  isoyear  = isoyear(key_date),
  isoweek  = isoweek(key_date),
  week     = week(key_date)
)

# week_date (Date) vs isoweek (bare number) as histogram x
imd_2000 %>%
  ggplot(aes(
    x = isoweek
  )) +
  geom_histogram(
    binwidth = 1,
    color    = "white"
  )

imd_2000 %>%
  ggplot(aes(
    x = week_date
  )) +
  geom_histogram(
    binwidth = 7,
    color    = "white"
  )

# What aweek would have handed you directly
date2week(
  as.Date(c("1999-12-31", "2000-01-01", "2000-01-02")),
  week_start = "Monday",
  floor_day  = TRUE,
  factor     = TRUE
)

# Labelling: monthly first
imd_2000 %>%
  ggplot(aes(
    x = week_date
  )) +
  geom_histogram(
    binwidth = 7,
    color    = "white"
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b"
  ) +
  theme_minimal()

# Labelling: ISO week/year, crowded
imd_2000 %>%
  ggplot(aes(
    x = week_date
  )) +
  geom_histogram(
    binwidth = 7,
    color    = "white"
  ) +
  scale_x_date(
    date_breaks = "2 weeks",
    date_labels = "%G-W%V"
  ) +
  theme_minimal()

# Labelling: rotated and legible
imd_2000 %>%
  ggplot(aes(
    x = week_date
  )) +
  geom_histogram(
    binwidth = 7,
    color    = "white"
  ) +
  scale_x_date(
    date_breaks = "2 weeks",
    date_labels = "%G-W%V"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

# Final plot
imd_2000 %>%
  ggplot(aes(
    x = week_date
  )) +
  geom_histogram(
    binwidth = 7,
    color    = "navy",
    fill     = "steelblue"
  ) +
  scale_x_date(
    date_breaks = "2 weeks",
    date_labels = "%G-W%V"
  ) +
  labs(
    x = "Epidemiological week",
    y = "Number of cases"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```
:::

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function() {
  var radiogroups = document.getElementsByClassName("webex-radiogroup");
  for (var i = 0; i < radiogroups.length; i++) {
    radiogroups[i].onchange = radiogroups_func;
  }
});
</script>
```
